The role of habitat quality in host metapopulations exposed to environmentally persistent pathogens. Or Identifying when environmentally persistent pathogens transform good habitat into ecological traps

### 2015-06-19 Manuscript/manuscript.tex
Small intro and methods edits

Many wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.  Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others (PCC: may need latin names, could include chytrid? I think it is also potentially noteworthy that these pathogens also seem to be particularly notorious in wildlife....therefore this is a big deal).  This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.  This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental longevity generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}).

@@

We expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}) (PCC: probably need a different reference for your first reference to "ecological trap").  In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).  Similarly, we expect that low quality patches will be less likely to develop pathogen reservoirs and may serve as refuges where susceptible hosts can escape infection.

@@

We assume that the host population size supported by a patch is proportional to its quality ($A_i$), and that disease-induced mortality reduces the size of infectious populations by a factor of $\nu$.  Building on the framework developed by \cite{Hanski2003} then, the rates at which patch \emph{i}, with quality $A_i$, is colonized by susceptible and infectious individuals, respectively, is given by its connectivity:
\begin{align}
C_{Si} &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(S_j)A_j^{\xi_{em}}\exp(-D d_{ij}),\\
C_{Ii} &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(I_j)(\nu A_j) ^{\xi_{em}}\exp(-D d_{ij}),
\label{connectivity}
\end{align}

PCC: not sure I understand why nothing is a function of time.

where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$  (PCC: not sure I would say $d_{ij}$ is a distance here when 2 paragraphs down you set $d_ij = 1$).  Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.

@@

The parameterization above assumes that all patches are equally accessible from all other patches (i.e. are separated by distance 1).  To examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighboring patches (in which case, average migration distance was adjusted to $D = 2$ to maintain roughly equivalent occupancy).

@@

To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat (PCC: while maintaining the same initial occupancy?).
These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality to roughly simulate alternative management actions that favor expansion (PCC: reword management actions that favor expansion) to include either additional high or low quality habitat (with qualities ranging from 0.76 to 1.77, and from 0.23 to 1.24 respectively).  
We then implemented a full-factorial design on each quality distribution in which the pathogen's disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).  In addition, pathogen longevity was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).

@@

A wide range of disease dynamics were observed over the range of pathogens and habitat quality distributions explored.  
Epidemic outcome was influenced most strongly by disease induced mortality, with low disease-induced mortality (i.e. high $\nu$) facilitating pandemic dynamics (in which no susceptible host populations persist), and very high disease-induced mortality (i.e. low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}). 
Pathogens with higher environmental longevities were generally able to persist (PCC: careful on word choice between persist and invade) and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics.  Increasing overall habitat quality facilitated pandemic spread for a larger range of pathogens (Fig~\ref{endemic}, a-c vs. d-f).
Metapopulations with a lattice structure exhibited qualitatively similar patterns, but with generally more limited pathogen spread and larger regions of endemic dynamics (Supplement).

For a wide range of parameter values, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions despite generally facilitating pathogen spread (red shaded regions in Fig~\ref{highvlow}).  (PCC: while true, one could also say: for the cases where the host and pathogen both persist there was a wide range of parameter values where low quality habitats supported more hosts.  Essentially, I'm intersecting Fig. 1 and 2, excluding the regions of the parameter space where the host or pathogen goes extinct (which is most of the red region in Fig 2). This is a very different sort of take home message. This is just a change of spin compared to how you phrase the paragraph.)
In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left corner of the panels in Fig~\ref{highvlow}). 
Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely (i.e. at a pandemic level) in both habitat quality distributions, high quality distributions again supported larger host populations than low quality distributions (top portion of panels in Fig~\ref{highvlow}).

@@

However, in the second scenario, where the pathogen's traits facilitate intermediate spread (i.e. at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics, with consequences for the overall host population size.
In particular, the ability of these pathogens to persist and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.

@@

This partitioning of pathogen dynamics into three regions based on whether high quality habitat is a net benefit or detriment to the host metapopulation echoes the work of (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
In particular, for a given pathogen, Hess (\cite{Hess1996}) similarly identified three regions of disease dynamics which determined whether increased movement had a positive or negative effect on host occupancy.  
In cases where the host movement rate was either low enough that the pathogen was unable to persist or high enough to cause a pandemic, Hess found that further increasing movement increased host occupancy, while at intermediate levels that resulted in endemic disease, increasing movement decreased host occupancy.
These three scenarios, distinguished in \cite{Hess1996} by the host movement rate, map to the the three scenarios we outline here based on pathogen traits (disease-induced mortality and direct and environmental transmission).
The conceptual similarity of these results suggests more generally that in metapopulations facing intermediate (i.e. endemic) pathogen spread, factors that would normally increase host stability and connectivity in the absence of disease (e.g. increased movement, high quality habitat) can actually decrease total host population size as a result of increased pathogen spread.

@@

\begin{figure}
\includegraphics[scale=0.5]{figure/figure_3.pdf}
\centering
\caption{Boxplots showing the range, over 100 replicate simulations, of susceptible (a), infectious (b), and total (c) host population sizes for high quality (red) and low quality (blue) habitat distributions.  Disease-induced mortality ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.3, respectively.  Colored points show medians, while vertical lines indicate the inner-quartile ranges, with horizontal lines indicating the minimum and maximum.  At low longevities, this pathogen is largely unable to persist and high quality habitat supports larger populations of susceptible hosts.  However, low quality habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}
\label{popsizes}
\end{figure}

>>>>

Many prominent and problematic wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.  Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\emph{Bacillus anthracis}, \cite{Dragon1995}), plague (\emph{Yersinia pestis}, \cite{Eisen2008}), and white nose syndrome (\emph{Geomyces destructans}, \cite{Lindner2011}), among others.  This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.  This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental longevity generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}).

@@

We expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Robertson2006}, \cite{Almberg2011}).  In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).  Similarly, we expect that low quality patches will be less likely to develop pathogen reservoirs and may serve as refuges where susceptible hosts can escape infection.

@@

We assume that the host population size supported by a patch is proportional to its quality ($A_i$), and that disease-induced mortality reduces the size of infectious populations by a factor of $\nu$.  Building on the framework developed by \cite{Hanski2003} then, the rates at which patch \emph{i}, with quality $A_i$, is colonized by susceptible and infectious individuals, respectively, is given by its connectivity:
\begin{align}
C_{Si}(t) &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(S, t)_jA_j^{\xi_{em}}\exp(-D d_{ij}),\\
C_{Ii}(t) &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(I, t)_j(\nu A_j) ^{\xi_{em}}\exp(-D d_{ij}),
\label{connectivity}
\end{align}

where $\phi(X, t)_j$ is an indicator function that is $1$ if patch $j$ is in state $X$ at time $t$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the effective distance between patches $i$ and $j$ .  Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.

@@

The parameterization above assumes that all patches are equally accessible from all other patches (i.e. are separated by effective distance 1).  To examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighboring patches (in which case, average migration distance was adjusted to $D = 2$ to maintain roughly equivalent occupancy).

@@

To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat.
These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality, with either the minimum or maximum shifted to simulate management actions that add either low or high quality habitat respectively (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77).  
We then implemented a full-factorial design on each quality distribution in which the pathogen's disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).  In addition, pathogen longevity was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).

@@

A wide range of disease dynamics were observed over the range of pathogens and habitat quality distributions explored.  
Epidemic outcome was influenced most strongly by disease induced mortality, with low disease-induced mortality (i.e. high $\nu$) facilitating pandemic dynamics (in which no susceptible host populations persist), and very high disease-induced mortality (i.e. low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}). 
Pathogens with higher environmental longevities were generally able to invade and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics.  Increasing overall habitat quality facilitated pandemic spread for a larger range of pathogens (Fig~\ref{endemic}, a-c vs. d-f).
Metapopulations with a lattice structure exhibited qualitatively similar patterns, but with generally more limited pathogen spread and larger regions of endemic dynamics (Supplement).

For a wide range of parameter values, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions despite generally facilitating pathogen spread (red shaded regions in Fig~\ref{highvlow}).  (PCC: while true, one could also say: for the cases where the host and pathogen both persist there was a wide range of parameter values where low quality habitats supported more hosts.  Essentially, I'm intersecting Fig. 1 and 2, excluding the regions of the parameter space where the host or pathogen goes extinct (which is most of the red region in Fig 2). This is a very different sort of take home message. This is just a change of spin compared to how you phrase the paragraph.)
In the case of pathogens that were unable to invade due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left corner of the panels in Fig~\ref{highvlow}). 
Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely (i.e. at a pandemic level) in both habitat quality distributions, high quality distributions again supported larger host populations than low quality distributions (top portion of panels in Fig~\ref{highvlow}).

@@

However, in the second scenario, where the pathogen's traits facilitate intermediate spread (i.e. at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics, with consequences for the overall host population size.
In particular, the ability of these pathogens to invade and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.

@@

This partitioning of pathogen dynamics into three regions based on whether high quality habitat is a net benefit or detriment to the host metapopulation echoes the work of (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
In particular, for a given pathogen, Hess (\cite{Hess1996}) similarly identified three regions of disease dynamics which determined whether increased movement had a positive or negative effect on host occupancy.  
In cases where the host movement rate was either low enough that the pathogen was unable to invade or high enough to cause a pandemic, Hess found that further increasing movement increased host occupancy, while at intermediate levels that resulted in endemic disease, increasing movement decreased host occupancy.
These three scenarios, distinguished in \cite{Hess1996} by the host movement rate, map to the the three scenarios we outline here based on pathogen traits (disease-induced mortality and direct and environmental transmission).
The conceptual similarity of these results suggests more generally that in metapopulations facing intermediate (i.e. endemic) pathogen spread, factors that would normally increase host stability and connectivity in the absence of disease (e.g. increased movement, high quality habitat) can actually decrease total host population size as a result of increased pathogen spread.

@@

\begin{figure}
\includegraphics[scale=0.5]{figure/figure_3.pdf}
\centering
\caption{Boxplots showing the range, over 100 replicate simulations, of susceptible (a), infectious (b), and total (c) host population sizes for high quality (red) and low quality (blue) habitat distributions.  Disease-induced mortality ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.3, respectively.  Colored points show medians, while vertical lines indicate the inner-quartile ranges, with horizontal lines indicating the minimum and maximum.  At low longevities, this pathogen is largely unable to invade and high quality habitat supports larger populations of susceptible hosts.  However, low quality habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}
\label{popsizes}
\end{figure}
